---
title: "Global Temperature Change"
subtitle: "A Swift Rise"
author: "Samarth Sulakhe"
institute: "University of Illinois at Chicago"
date: "05-03-2026"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
library(tidyverse)
library(palmerpenguins)
library(knitr)
library(xaringanthemer)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(fig.retina = 3, dpi = 300, fig.width = 6, fig.asp = 0.618, out.width = "80%")
```

```{r load-data, include=FALSE}
GlobalTemperatures <- read_csv("../data/GlobalTemperatures.csv")
GlobalLandTemperaturesByCountry <- read_csv("../data/GlobalLandTemperaturesByCountry.csv")
```

```{r title-background, include=FALSE}
style_xaringan(
  title_slide_background_color = "red"
)
```



class: center, middle

### Research Question  
How have global and country-level temperatures changed over time, and are these trends statistically significant?

---

class: inverse, center, middle

# Data

---

# Data Overview

- Two datasets:
  - **GlobalTemperatures**
  - **GlobalLandTemperaturesByCountry**

- Observations include:
  - Date (dt)
  - Temperature values (°C)
  - Country (for country-level data)

- Each row = one temperature observation at a specific time and location

---

# Key Variables

- **AverageTemperature / LandAverageTemperature** → response (Y)  

- **Year (derived from date)** → predictor (X)  

- **Country** → comparison groups  

---

class: inverse, center, middle

# Exploratory Analysis

---

# Global Temperature Trend

```{r global-trend, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7,fig.height=6}
library(dplyr)
library(ggplot2)

GlobalTemperatures$Year <- as.numeric(format(as.Date(GlobalTemperatures$dt), "%Y"))

avg_temp_year <- GlobalTemperatures %>%
  group_by(Year) %>%
  summarize(mean_temp = mean(LandAverageTemperature, na.rm = TRUE)) %>%
  filter(Year >= 1850)

ggplot(avg_temp_year, aes(x = Year, y = mean_temp)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Average Global Land Temperature Over Time",
       x = "Year",
       y = "Temperature (°C)")
```

---

# Country Comparison

```{r global-land, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7,fig.height=6}
GlobalLandTemperaturesByCountry$Year <- as.numeric(format(as.Date(GlobalLandTemperaturesByCountry$dt), "%Y"))

subset_data <- GlobalLandTemperaturesByCountry %>%
  filter(Country %in% c("United States", "India", "China"))

plot_data <- subset_data %>%
  group_by(Year, Country) %>%
  summarize(mean_temp = mean(AverageTemperature, na.rm = TRUE)) %>%
  filter(Year >= 1850)

ggplot(plot_data, aes(x = Year, y = mean_temp, color = Country)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Temperature Trends by Country",
       x = "Year",
       y = "Temperature (°C)")
```

---

# Key Takeaways

- **Global Temperature Trends**

  - Clear upward trend throughout time
  
  - Strong increase after the 20th century
  
- **Country Comparison**

  - All countries show general increasing trends
  
  - Solidifies the "global warming" argument


---

class: inverse, center, middle

# Statistical Analysis

---

# Linear Regression (Global)

```{r global-lin-reg, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
model_global <- lm(mean_temp ~ Year, data = avg_temp_year)
summary(model_global)
```

---
```{r global-lin-reg-cont, echo=FALSE, message=FALSE, warning=FALSE}
model_global <- lm(mean_temp ~ Year, data = avg_temp_year)
summary(model_global)
```
  - Tests if temp. increases through time
  
  - If the p-value is less than 0.05, this means it is a statistically significant trend

---

# 30 Year Climate Norms

```{r climate-norm, echo=FALSE, message=FALSE, warning=FALSE}
avg_temp_year <- avg_temp_year %>%
  mutate(block_30yr = floor(Year / 30) * 30)

climate_normals <- avg_temp_year %>%
  group_by(block_30yr) %>%
  summarize(mean_temp = mean(mean_temp, na.rm = TRUE))

ggplot(climate_normals, aes(x = block_30yr, y = mean_temp)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "30-Year Climate Normals",
       x = "Year (30-year blocks)",
       y = "Temperature (°C)")
```

---

# Bootstrapping

```{r climate-bootstrap, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(123)

boot_ci <- avg_temp_year %>%
  group_by(block_30yr) %>%
  summarize(
    boot_means = list(replicate(300, {
      samp <- sample(mean_temp, replace = TRUE)
      mean(samp)
    }))
  ) %>%
  rowwise() %>%
  mutate(
    lower = quantile(unlist(boot_means), 0.025),
    upper = quantile(unlist(boot_means), 0.975),
    mean = mean(unlist(boot_means))
  )

ggplot(boot_ci, aes(x = block_30yr, y = mean)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 5) +
  theme_minimal() +
  labs(title = "30-Year Temperature with 95% CI",
       x = "Year (30-year blocks)",
       y = "Temperature (°C)")
```

---

# Key Takeaways

- **30 Year Climate Norms**

  - Chart smooths out short-term fluctuation
  
  - Highlights long-term warming trends
  

- **Bootstrapping**

  - Displays a variability in estimates
  
  - With non overlapping intervals, we attain significant differences

---

class: inverse, center, middle

# Conclusion

---

#Key Findings

- Global temperatures show a clear upward trend

- Regression confirms trend is statistically significant

- 30-year averages show consistent increase

- Bootstrapped confidence intervals indicate significant differences between periods

---

class: center, middle

# Thank You!
